---
title: "Tests of life-history trade-offs"
author: "Elizabeth Ostrowski"
date: "08/09/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

Load packages
```{r}
library(tidyverse)
library(cowplot)
library(glmmTMB)
library(emmeans)

# X <- df[, c("mean_spef","mean_germ")]
# md <- mahalanobis(X, colMeans(X), cov(X))
# cutoff <- qchisq(0.975, df=2)   # ≈ 5.99
# df$outlier <- md > cutoff
# 
# library(ggplot2)
# 
# ggplot(df, aes(x=mean_germ, y=mean_spef))+
#  geom_point(aes(color=outlier), size=3)+
#  geom_smooth(data=df, method="lm", se=FALSE, color="black")+
#  geom_smooth(data=df_no_out, method="lm", se=FALSE, color="red", linetype="dashed")+
#  theme_bw(base_size=14)
# 
# X <- df[, c("mean_spef","mean_germ")]
# md <- mahalanobis(X, colMeans(X), cov(X))
# cutoff <- qchisq(0.975, df=2)   # ≈ 5.99
# df$outlier <- md > cutoff


```

Plotting
```{r}
strains <- c( AX4 = "rAX4", CD1 = "rCD1",CD2 = "rCD2", CD3 = "rCD3", CD4 = "rCD4", Control = "WT Control") 
colours <- c(rCD1 = "#C06C63",rCD2 = "#AFC377", rCD3 = "#71B8CF", rCD4 = "#E9A864", rAX4 = "black") 
```

Read and clean doubling time (dt) data
```{r}

#Read in data
dt = read.table("datafiles/doubling_times.txt", header=T, sep="\t")

dt$evol_group <- recode_factor(dt$evol_group, CD1="rCD1", CD2="rCD2", CD3="rCD3", CD4="rCD4", AX4 = "rAX4") # rename & reorder levels

# Set a reference line for the ancestor
dt_anc_value <-  as.numeric(dt %>% filter(strain=="AX4_GFP") %>% summarise(mean_dt=mean(dt, na.rm=T)))

# Average across replicates (= dt_sum)
dt_sum = dt %>% group_by(strain, evol_group) %>% summarise(mean_dt = mean(dt, na.rm=T)) 

# For plotting, we will filter to remove non-evolved strains
dt_for_plot = dt_sum %>% filter(., grepl("r",strain)) # strains that start with "r" (=evolved)
```

Read and summarise sporulation efficiency (spef) data
```{r}

resist = read.table('datafiles/evolved_isolates_resistance.txt', header=T, sep='\t', stringsAsFactors = F,na.strings = "NA") 

# Columns show the clonal sporulation efficiency of each strain, but that strain was mixed with multiple strains in each block, so we want just the unique values, representing clonal sporulation of that strain on that day. Need to do separately for the unlabeled ("unl") and labeled ("gfp") strains & then merge
spef_unl_unique = resist %>% group_by(strain_unl) %>% dplyr::select(strain_unl, spef_unl, exp_date) %>% unique
spef_gfp_unique = resist %>% group_by(strain_gfp) %>% dplyr::select(strain_gfp, spef_gfp, exp_date, evol_partner) %>% unique

names(spef_unl_unique) = c('strain', 'spef', 'expt_date')
names(spef_gfp_unique) = c('strain', 'spef', 'expt_date', 'evol_partner')
spef_raw = rbind(spef_gfp_unique, spef_unl_unique) 

spef = spef_raw %>% group_by(strain) %>% summarise(mean_spef = mean(spef, na.rm=T)) 

# Recode the evol_partner column: 
spef_sum = spef %>%
  mutate(evol_group =
    case_when(
        startsWith(strain, "rCD1") ~ "rCD1",
        startsWith(strain, "rCD2") ~ "rCD2",
        startsWith(strain, "rCD3") ~ "rCD3",
        startsWith(strain, "rCD4") ~ "rCD4",
        startsWith(strain, "rAX4") ~ "rAX4",
        strain=='AX4' ~ "wt",
        strain=="AX4_G" ~ "anc",
        startsWith(strain, "CD") ~ "cheater",
                                  )
  )
                                           
spef_anc_value = subset(spef_sum, evol_group=="anc")$mean_spef
spef_for_plot = subset(spef_sum, evol_group!="anc" & evol_group!="wt" & evol_group!="cheater")
spef_for_plot$evol_group <- recode_factor(spef_for_plot$evol_group, CD1="rCD1", CD2="rCD2", CD3="rCD3", CD4="rCD4", AX4 = "rAX4") # rename & reorder levels
```

Read and summarise germination efficiency (germ) data
```{r}

germ_raw = read.table("datafiles/germination.txt", sep='\t', header=T)
names(germ_raw)[4] = c('evol_group') # rename columns
germ_raw$prct_germ <- (germ_raw$spores.start - germ_raw$spores.end)/germ_raw$spores.start

germ = germ_raw %>% group_by(strain, evol_group) %>% summarise(cnt=n(), mean_germ = mean(prct_germ, na.rm=T))

germ_sum = germ %>%
  mutate(evol_group =
    case_when(
        startsWith(strain, "rCD1") ~ "rCD1",
        startsWith(strain, "rCD2") ~ "rCD2",
        startsWith(strain, "rCD3") ~ "rCD3",
        startsWith(strain, "rCD4") ~ "rCD4",
        startsWith(strain, "rAX4") ~ "rAX4",
        strain=='AX4' ~ "wt",
        strain=="AX4_G" ~ "anc",
        startsWith(strain, "CD") ~ "cheater",
                                  )
  )

germ_anc_value = germ_sum[germ_sum$strain=="AX4",]$mean_germ
germ_for_plot = germ_sum %>% filter(., evol_group != "anc" & evol_group != "wt" & evol_group != "cheater")
germ_for_plot$evol_group <- recode_factor(germ_for_plot$evol_group, CD1="rCD1", CD2="rCD2", CD3="rCD3", CD4="rCD4", AX4 = "rAX4") # rename & reorder levels

```

Figure 7A, doubling time 
```{r}

p.dt<-ggplot(data=dt_for_plot) + 
  geom_boxplot(aes(x=evol_group, y=mean_dt,  col='white'), outlier.shape=NA, show.legend=F) + 
  geom_jitter (aes(x=evol_group, y=mean_dt, col=evol_group), width=0.1, size=2.5, alpha=0.7, show.legend=F) + 
  theme_cowplot() + ylim(5,16) +
  scale_fill_manual(values=colours) + 
  labs(y = "\nDoubling Time (hours)", x="Evolution Group") +
  scale_colour_manual(values=colours) + 
  geom_hline(yintercept = dt_anc_value, lty=2) + 
  theme(axis.title = element_text(size = 12), 
                     axis.text = element_text(size = 12),
                     panel.grid.major = element_blank(), 
                     panel.grid.minor = element_blank(),
                     axis.text.x = element_text(angle = 45, hjust = 1),
                     axis.line = element_line(colour = "black")) 

```

Figure 7B, sporulation efficiency ('spef')
```{r}

p.spef<-ggplot(data=spef_for_plot) + 
  geom_boxplot(aes(x=evol_group, y=mean_spef, col='white'), outlier.shape = NA, show.legend=F) + 
  geom_jitter(aes(x=evol_group, y=mean_spef, col=evol_group), width=0.1, size=2.5, alpha=0.7, show.legend=F) + 
  theme_cowplot()  +
  labs(y = "Sporulation Efficency", x=" Evolution Group") + 
  scale_colour_manual(values=colours) + 
  geom_hline(yintercept = spef_anc_value, lty=2) + 
  theme(axis.title = element_text(size = 12), 
                     axis.text = element_text(size = 12),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     axis.text.x = element_text(angle = 45, hjust = 1),
                     axis.line = element_line(colour = "black")) 

```

Figure 7C, Germination efficiency
```{r}

p.germ<-ggplot(data=germ_for_plot) + 
  geom_boxplot(aes(x=evol_group, y=mean_germ, col='white'), outlier.shape=NA, show.legend=F) +
  geom_jitter (aes(x=evol_group, y=mean_germ, col=evol_group), width=0.1, size=2.5, alpha=0.7, show.legend=F) + 
  theme_cowplot() + ylim(0.7,1)+
  labs(y="Germination Efficiency", x=" Evolution Group") +
  scale_colour_manual(values=colours) + 
  geom_hline(yintercept = germ_anc_value, lty=2) + 
  theme(axis.title = element_text(size = 12), 
                     axis.text = element_text(size = 12),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     axis.text.x = element_text(angle = 45, hjust = 1),
                     axis.line = element_line(colour = "black")) 
```

Figure 7D, Correlation between sporulation and germination
```{r}

data_1 = merge(spef_sum, germ_sum, by="strain")
data_2 = merge(data_1, dt_sum, by="strain")
data_all = data_2 %>% select(strain, evol_group.x, mean_spef, mean_germ, mean_dt) %>% rename(evol_group=evol_group.x)
data_evolved_only = data_all %>% filter(., grepl("r", strain))

# Correlations among all strains
cor.test(data_all[,3], data_all[,4]) # corr betw spef and germ: r=-0.35, p=0.04
cor.test(data_all[,3], data_all[,5]) # corr betw spef and dt: r=-0.04, p=0.82
cor.test(data_all[,4], data_all[,5]) # corr betw germ and dt: r=-0.07, p=0.68

# Correlations among the evolved strains only
cor.test(data_evolved_only[,3], data_evolved_only[,4]) # corr betw spef and germ: r=-0.24, p=0.21
cor.test(data_evolved_only[,3], data_evolved_only[,5]) # corr betw spef and dt: r=-0.01, p=0.94
cor.test(data_evolved_only[,4], data_evolved_only[,5]) # corr betw germ and dt: r=-0.09, p=0.63

# Correlations among cheater-evolved strains only
data_evo_rm_rAX4 = subset(data_evolved_only, data_evolved_only$evol_group!='rAX4')
cor.test(data_evo_rm_rAX4[,3], data_evo_rm_rAX4[,4]) # corr betw spef and germ: r=-0.399, df=22, p=0.053
cor.test(data_evo_rm_rAX4[,3], data_evo_rm_rAX4[,5]) # corr betw spef and dt: r=0.003, df=22, p=0.988
cor.test(data_evo_rm_rAX4[,4], data_evo_rm_rAX4[,5]) # corr betw germ and dt: r=0.018, df=22, p=0.935


colours <- c(rCD1 = "#C06C63",rCD2 = "#AFC377", rCD3 = "#71B8CF", rCD4 = "#E9A864", rAX4 = "black", wt="dodgerblue2", "cheater" = "firebrick2") 

ggplot(data=data_all, aes(x=mean_spef, y=mean_germ,col=evol_group)) + geom_smooth(method='lm',col="black", lwd=0.5,fill="grey") + geom_point() + xlab('Sporulation Efficiency') + ylab('Germination Efficiency') + scale_colour_manual(values=colours) + theme_cowplot() + theme(legend.position = c(0.8,0.85),legend.key.size = unit(0.3,"cm"), legend.text = element_text(size=6), legend.title=element_blank(),legend.spacing.y = unit(0, 'cm')) + guides(fill = guide_legend(byrow = TRUE)) + labs(col=none) + ylim(c(0.7,1))

p.corr<-ggplot(data=data_evo_rm_rAX4, aes(x=mean_spef, y=mean_germ,col=evol_group)) + geom_smooth(method='lm',col="black", lwd=0.5,fill="grey") + geom_point() + xlab('Sporulation Efficiency') + ylab('Germination Efficiency') + scale_colour_manual(values=colours) + theme_cowplot() + theme(legend.position = c(0.8,0.85),legend.key.size = unit(0.3,"cm"), legend.text = element_text(size=6), legend.title=element_blank(),legend.spacing.y = unit(0, 'cm')) + guides(fill = guide_legend(byrow = TRUE)) + labs(col=none) + ylim(c(0.7,1))

```

```{r}


p.top = plot_grid(p.dt, p.spef, labels=c("A", "B"), align="h", axis="b")
p.bottom = plot_grid(p.germ, p.corr, align="h", axis="b", labels=c("C", "D"))
p.all = plot_grid(p.top,p.bottom, ncol=1)
save_plot("figures/Figure_7.pdf", p.all,base_height=7, base_width=7)
```

Calculate mean cheating ability
```{r}

mean_cheat = resist %>% group_by(strain_gfp, strain_unl) %>% summarise(mean_spores_gfp = mean(spores_prct_gfp)) %>% group_by(strain_gfp) %>% summarise(gd_mean_spores_gfp = mean(mean_spores_gfp))

d=merge(data_all, mean_cheat, by.x="strain", by.y="strain_gfp")
names(d)[6] = "mean_cheat"
cor.test(d$mean_spef,d$mean_cheat)  # r = 0.81, p<0.0001
cor.test(d$mean_cheat, d$mean_germ) # r = -0.22, p=0.25 
cor.test(d$mean_cheat, d$mean_dt)   # r = 0.25, p=0.18
cor.test(d$mean_spef, d$mean_germ)  # r = -0.24, p=0.21 (same as above)

```

Statistics - germination
```{r}
germ_raw_evol_only = germ_raw %>% filter(., grepl("r",strain)) 

m.germ = glmmTMB(prct_germ ~ (1|strain) + evol_group + (1|block), data=germ_raw_evol_only, family=beta_family(link="logit"))

summary(m.germ)
drop1(m.germ, test="Chisq") # group is significant
emmeans(m.germ, specs = pairwise ~ evol_group)


# Drop strain
m1.germ = glmmTMB(prct_germ ~ (1|strain) + evol_group + (1|block), data=germ_raw_evol_only, family=beta_family(link="logit"))
anova(m.germ, m1.germ, test="Chisq") # strain is NS

# For posthoc testing that rAX4 is different from rCDs
germ_raw_evol_only = germ_raw_evol_only %>% mutate(mixtype = case_when(
  evol_group == "AX4" ~ "no_cheater", 
              evol_group !="AX4" ~ "cheater"))
m.germ = glmmTMB(prct_germ ~ (1|strain) + mixtype + (1|block), data=germ_raw_evol_only, family=beta_family(link="logit"))

emmeans(m.germ, specs = trt.vs.ctrl ~ mixtype)

```

Statistics - sporulation efficiency
```{r}

spef_raw_evol_only = spef_raw %>% filter(., grepl("r",strain)) 

m.spef = glmmTMB(spef ~ (1|strain) + evol_partner + (1|expt_date), data=spef_raw_evol_only)

summary(m.spef)
drop1(m.spef, test="Chisq") # evol_partner is NS

# Drop strain
m1.spef = glmmTMB(spef ~ evol_partner + (1|expt_date), data=spef_raw_evol_only)
anova(m.spef, m1.spef, test="Chisq") # strain is significant
emmeans(m.spef, specs = pairwise ~ evol_partner) # CD3 is different from CD4; all else NS

```

Statistics - doubling time 
```{r}

dt_raw_evol_only = dt %>% filter(., grepl("r",strain)) 

m.dt = glmmTMB(dt ~ (1|strain) + evol_group + (1|rep), data=dt_raw_evol_only)

summary(m.dt)
drop1(m.dt, test="Chisq") # evol_group is NS

# Drop strain
m1.dt = glmmTMB(dt ~ evol_group + (1|rep), data=dt_raw_evol_only)
anova(m.dt, m1.dt, test="Chisq") # strain is NS
emmeans(m.dt, specs = pairwise ~ evol_group) # all strains NS

```

